WHITE WINE QUALITY "SRIJON SARKAR"
========================================================

```{r global_options, include=FALSE}

knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(dplyr)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv('whitewineq.csv')
```

> In this dataset, we will be exploring the different features of white wine. We
shall compare the different attributes with the quality of the wine itself and 
try to learn which features are most correlated with it.

# Univariate Plots Section


```{r Univariate_Plots}
str(df)
```

>We have got 4898 samples in our dataset and each sample has 13 observations.
>One thing we notice upon seeing the structure of the dataset is that the output variable(quality) , even though it always takes discrete values(0-10), is not a 
factor variable. It makes sense to convert it into a factor variable. 


```{r }
df$quality <- factor(df$quality)
str(df$quality)
```

```{r }
summary(df)
```

>Now,let us plot a histogram of the different qualities of the white wine.

```{r echo=FALSE}
ggplot(aes(quality), data=df) + geom_bar(color='blue', fill='blue')

```

>What we observe here is that although wine quality ranges from 0-10, we only
have qualities in 3-9 in our dataset. Also, count of quality 3 and 9 wine is 
very low. The highest count of white wine is of quality 6.

>Now let us plot a histogram of the pH levels of all the samples.

```{r echo=FALSE}
ggplot(aes(pH), data=df) + geom_histogram(color='red', fill='blue', binwidth=0.03)

```
```{r }
summary(df$pH)
```


>As the plot visually signifies, most of the pH values lie around the point 
3.18. The summary on pH attribute shows that the mean is at 3.18, while the 
median is at 3.188. 

>Now lets look at the distribution of density in the different wine samples. 

```{r echo=FALSE}
ggplot(aes(x=density), data=df) + geom_histogram(color='red',
    fill='orange', binwidth=0.001)

```

>We can get a decent idea of the distribution here, but most values are centered
just below 1, and rest of the plot has outliers. 

```{r }
summary(df$density)
```


>As we see here, the minimum value of density is 0.987 and maximum value is 1.039, 
which is very far away. Let us plot only those points above 0.98 and below 1.01. 

```{r echo=FALSE}
ggplot(aes(x=density), data=df) + geom_histogram(color='red',
    fill='orange', binwidth=0.001) + xlim(0.98, 1.01)

```

>This plot is a better visualisation. We notice that most peaks are just above
0.99, and all counts above 200 are between 0.99 and 1.


Let us plot the histogram of alcohol variable and see its distribution.

```{r echo=FALSE}
ggplot(aes(alcohol), data=df) + geom_histogram(color='green', binwidth=0.5)
```

>This histogram tells us that a large no. of our samples have alcohol% just 
below 10. All the alcohol% values lie between 8 and 14. Let us plot a boxplot 
and compare the values with different qualities of white wine.

```{r echo=FALSE}
# FIXED ACIDITY
ggplot(aes(fixed.acidity), data=df) + geom_histogram(color='green',binwidth=0.35)+
    xlim(4,11)
```

>Most white wine samples in our dataset have fixed acidity levels around 6.5 g/dm^3.

```{r echo=FALSE}
#volatile acidity
ggplot(aes(volatile.acidity), data=df) + geom_histogram(color='green',binwidth=0.05) +
    xlim(0, 0.8)
  
```

>All our samples have volatile acidity values below 0.8 g / dm^3, and maximum samples
have between 0.2 and 0.3 values.

```{r echo=FALSE}
#total sulfur dioxide
ggplot(aes(total.sulfur.dioxide), data=df) + geom_histogram(color='green',
       binwidth=15) + xlim(0, 285)
```

>It refers to the total sulfur dioxide contents of white wine, including free and
bound forms. It is centered around 120 mg / dm^3.


```{r echo=FALSE}
# free sulfur dioxide
ggplot(aes(free.sulfur.dioxide), data=df) + geom_histogram(color='green',
          binwidth=8) + xlim(0, 120)
```

>Most free sulfur dioxide values are around the 25 mg / dm^3 mark.

```{r echo=FALSE}
#residual sugar
ggplot(aes(residual.sugar), data=df) + geom_histogram(color='green') +
      xlim(0,25)
```

>The residual sugar is measured in g / dm^3 , and we can see that maximum samples
in our dataset have residual sugar amount below 2.5 .

```{r echo=FALSE}
#chlorides
ggplot(aes(chlorides), data=df) + geom_histogram(color='green',binwidth=0.01) + 
      xlim(0, 0.15)
```

>Chlorides is the amount of salt in wine, measured in g/ dm^3, and most of our 
samples have below 0.05

# Univariate Analysis

### What is the structure of your dataset?

>The white wine quality dataset has 4898 no. of samples, with each sample 
associated with 12 features( +1 which is indices). The output variable is the 
quality of each sample which can ideally have a number from 0 to 10, however our
dataset only has samples qualities lying between 3 and 9. 
The maximum no. of samples belong to quality 6.

### What is/are the main feature(s) of interest in your dataset?

>I am interested mostly in the quality factor, and the features which help in 
driving that quality upwards, that is which variables are highly correlated with
the quality factor.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

>The alcohol% , pH and density seem to be important variables with respect to 
determining the quality, and hence will concentrate on them. But will look for 
correlations in other variables too. 

### Did you create any new variables from existing variables in the dataset?

>No, I did not create any new variable.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \ of the data? If so, why did you do this?

>Yes, I converted the type of data for the quality factor. Since it took a 
discrete value from the set(0,10), I converted it into a factor variable. That's 
the only change I did in the dataset.


# Bivariate Plots Section

> Let us first plot fixed.acidity vs volatile.acidity, and check if there is any
correlation between the two.

```{r echo=FALSE, Bivariate_Plots}
ggplot(aes(fixed.acidity, volatile.acidity), data=df) +
  geom_point(alpha=1/5, color='blue')
```

>Based on this plot, it doesn't seem that there is any correlation in between 
the variables. Let us find the correlation coeficient of the same.

```{r echo=FALSE}
 cor.test(df$fixed.acidity, df$volatile.acidity)
```

>The value of correlation coefficient is too low to be considered significant. 
Hence we can conclude that there isn't any correlation between them.


>Let us see if the density changes along with the quality of wine. Let us make a
box plot for the different wine qualities. 

```{r }
ggplot(aes(x=quality, y=density), data=df) + geom_boxplot(color='orange') +  ylim(0.98, 1.01)

```

>We can see that the median value of density of the white wine generally gets 
lower as the quality of the wine increases. Although there are quite a few 
outliers of lower quality which also show similar density as higher quality wines. 
Let us find the actual numerical differences in the median values.

>Density stats grouped by different qualities.

```{r echo=FALSE}
by( df$density, df$quality, summary)
```

>The differences are very small, but of course there. The median value of density 
drops from 0.9944 to 0.9903 from quality 3 to 9. The maximum value of density for 
quality 9 is also below 1, unlike other qualities. This signifies that higher 
quality wines have lower densities.

>Let us see if there is a relation between alcohol % and quality. 

```{r echo=FALSE}
ggplot(aes(quality,  alcohol), data=df) + geom_boxplot(color='red')

```

Summary of different alcohol% based on quality:

```{r echo=FALSE}
by(df$alcohol, df$quality, summary)
```

>This boxplot and summary shows that the alcohol % increases as the quality 
increases. There is a steady increase in alcohol % from level 5 to 9. That is, 
higher quality wines have higher alcohol contents. 

>let us plot a boxplot of different pH values of the different qualities and find
the differences, if any.

```{r echo=FALSE}
ggplot( aes(x=quality, y=pH), data=df) + geom_boxplot(color='red', fill='grey')

```

> It is somewhat similar to the above plot. Just like how alcohol% decreases 
from quality level 3 to 5 and then increases, median pH values follows the same
trend. This means that there could be some kind of correlation between pH and 
alcohol%. Let us plot these 2 variables on a scatter plot and see if any 
correlation exists.

```{r echo=FALSE}
ggplot(aes(pH, alcohol), data=df) + geom_point(color='orange')
```

> The plot seems very spread out and dispersed. The correlation coefficient of 
these features must be quite low. It seems unlikely that any correlation exists;
however we'll find the pearson's correlation coefficient anyway.

```{r echo=FALSE}
cor.test(df$pH, df$alcohol)
```

>The correlation is quite low in these two variables; so there isn't any 
causation between the two, but these 2 variables do cause another variable(quality). 

>Apart from that, we saw that higher quality wines have higher pH values(more
basic). This could mean that higher quality wines have lesser acidic properties.
Let us plot a box plot of fixed acidity against the qualities.

```{r echo=FALSE}
ggplot(aes(x=quality,y=fixed.acidity),data=df) +
  geom_boxplot(color='darkgreen') + ylim(4.5,10.5)
```

>There is a very small but appreciatable decrease in the fixed acidity levels, 
as the quality increases. However, it is not significant enough. 

>Now lets plot free.sulfur.dioxide vs total.sulfur.dioxide . Without seeing the 
plot, we know that there must be a correlation between the two, as total sulfur 
dioxide takes account of the free sulfur dioxide already. 

```{r echo=FALSE }
ggplot(aes(free.sulfur.dioxide, total.sulfur.dioxide), data=df) +
 geom_point(color='red', alpha=0.25) + xlim(0,150) + ylim(0,300)
```

>Just as we felt, the plot does signify a strong correlation between the 2 
features. Lets find the correlation coeficient between them.

```{r echo=FALSE}
cor.test(df$free.sulfur.dioxide, df$total.sulfur.dioxide)
```

>We can see that the correlation value is quite high. Hence we can conclude that
there is a strong correlation between the two, and a causation of course, as total
sulfur dioxides accounts for free and bound forms of SO2.


>Let us find the correlation between density and residual sugar, if any.

```{r echo=FALSE}
ggplot(aes(residual.sugar,density),data=df) + 
  geom_point(alpha=1/10, color='navy')  +  xlim(0,30)
```

>It is pretty evident that there is a strong correlation between these 2 features.
Let us find the correlation coefficient of the same.

```{r echo=FALSE}
cor.test(df$density, df$residual.sugar)
```

>The correlation coefficient here is very high, which shows a strong correlation
between density and residual sugar features. 

```{r echo=FALSE}
ggplot(aes(quality,chlorides),data=df) + geom_boxplot() +
  ylim(0.0,0.1)
```

>There is a significant decrease in the chlorides level of white wine with the 
increase in quality. 

>Let us plot density vs chlorides, as we had seen that like chlorides, density 
had also decreased with quality. 

```{r, echo=FALSE}
ggplot(aes(density,chlorides),data=df) + geom_jitter(alpha=1/10, color='blue')  + xlim(0.987,1.003)
```

> We can see there is a region which is slightly overplotted, but also seems 
correlated. Let us find the correlation factor of the same.

```{r echo=FALSE}
cor.test(x=df$density,y=df$chlorides)
```

>The correlation factor here is not significant of course, and hence we cannot 
conclude that the variables are correlated.

>Let us plot density vs alcohol and test its correlation.

```{r echo=FALSE}
ggplot(aes(alcohol,density),data=df)+  geom_jitter(color='red', alpha=0.3)
```

>It seems as if they are correlated very closely, let us find correlation factor
to confirm this.

```{r echo=FALSE}
cor.test(df$alcohol,df$density)
```

>As we can see the correlation factor is a negative high value, which means that
they are strongely negatively correlated. Thus, we infer that as the alcohol% 
increases in the wine sample, its density decreases. 

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

> It is seen that higher qualities of white wine have got lower densities. Also, 
high qualities of wine also have higher values of pH and alcohol% than the other 
levels.
>2 unusual distributions are there in pH and alcohol % values however. Generally
both increases with the quality, however quality 3 and 4 have higher pH and alcohol% 
values than level 5 and hence going against the trend.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

>The residual sugar and density features have a very high degree of correlation.
Alcohol% and density also had a strong correlation, but negatively. Also, the 
fixed SO2 and free SO2 levels in white wine are highly correlated too., but it 
is somewhat expected so.


### What was the strongest relationship you found?

>The highest degree of correlation is found between residual sugar and density 
values. 

# Multivariate Plots Section


>We've seen that higher quality of wine have got lower amounts of densities, 
while having pH and alcohol%. So let us plot alcohol vs pH for the different 
qualities first.


```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(alcohol,pH),data=df)+  geom_point(aes(color=quality)) + 
    scale_color_brewer(palette="RdYlBu")
```

>Most blue spots, which correspond to high quality, are found at the right side
of the scatter plot, which means high % alcohol. However, no visible trend is present corresponding to the pH values for lower or higher quality wines.

>Now let us plot density vs alcohol vs quality


```{r echo=FALSE}
ggplot(aes(alcohol,density),data=df)  +
  geom_point(aes(color=quality), alpha=0.5) + ylim(0.988,1.005) + 
  scale_color_brewer(palette="Blues")
``` 

>This plot is spread out similar to the above plot. The lighter spots are 
concentrated to the left, and darker ones in the right of the scatter plot. Blue
spots are to the right and below spots meaning high alcohol and lower densities,
so high quality wines have higher qualities of wine and lower densities.

```{r echo=FALSE}
ggplot(aes(density,pH),data=df)  +
  geom_point(aes(color=quality), alpha=0.5) + xlim(0.988,1.005) + scale_color_brewer(palette="Greens")
```

>It is somewhat hard to conclude something concrete from this plot, as the light
green spots are all over the plot. However, we can see  most darker spots are to
the left of the plot, signifying lower density values for high quality wines.


```{r echo=FALSE}
ggplot(aes(residual.sugar,alcohol,color=density),data=df)   +  geom_point() +
     scale_color_gradient(low="blue", high="red") + xlim(0,22)
```

>In this plot, we see that pure blue colour spots are present only at the top 
left region, which means lowest values of density corresponds to high percentage
of alcohol and low amounts of residual sugar. As the amount of sugar increases 
and alcohol decreases, the spot colour changes towards red, which means high density.

>It seems as if residual sugar and alcohol are good estimators for the density 
variable. Let's try to fit a linear model to this.

```{r echo=FALSE}
m1 <- lm(density ~ alcohol + residual.sugar, data = df)
summary(m1)
```

> Just the alcohol and residual sugar variables account for 90% of variance in 
density variable.


```{r echo=FALSE}
ggplot(aes(fixed.acidity,volatile.acidity,color=pH),data=df)   +  geom_point() + scale_color_gradient(low="blue", high="red")

```

> We see here that the red spots are concentrated to the left and lower parts of
the scatter plot, which means higher pH values have lower fixed and volatile 
acidity values. This is expected, as high pH values correspond to lesser acidic 
properties.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

>Not exactly strengthened, as most scatter plots which took on colours corresponding 
to quality levels were not linearly seperable. Generally it was visible that high 
quality wines have lower density, higher alcohol% and higher pH values. 

### Were there any interesting or surprising interactions between features?

>Residual sugar and alcohol are major factors of density of a particular sample.
A clear transition is seen in the scatter plot.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

>Yes, I created a linear model starting from the residual sugar and the alcohol% to the 
density.

>The variables in the linear model account for 90.7% of the variance in the density
of white wine. I have used only 2 variables in this model; there must be more
variables which affect the density variable.



------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, Plot_One}

ggplot(aes(quality,  alcohol), data=df) + geom_boxplot(color='red',fill='grey')+
   ggtitle("Quality vs Alcohol%") + ylab("alcohol (% by volume)")
```

### Description One
What we observe here is the variation in alcohol% in the different quality levels
of white wine. While there is a slight decrease from quality 3 to 5, from there
it increases significantly to the highest quality.


### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(residual.sugar,alcohol,color=density),data=df)   +  geom_point() +
     scale_color_gradient(low="blue", high="red") + xlim(0,22) + 
    labs(title = "Residual sugar vs Alcohol\n", x = "Residual sugar g/dm^3",
       y = "Alcohol % by volume", color = "Density (g / cm^3)\n")
```

### Description Two
This plot is significant as it shows the variation in density as a function of 
alcohol and residual sugar. Fitting these to a linear model gave us one in which
alcohol and residual sugar accounted for 90% variation in density.


### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(alcohol,pH),data=df)+  geom_point(aes(color=quality)) + 
    scale_color_brewer(palette="RdYlBu") + ggtitle("Alcohol vs pH") + 
  xlab("Alcohol (% by volume)") + ylab("pH")
```

### Description Three

This plot shows the variation of quality with pH and alcohol% . Just as we had 
seen before, higher quality wines(8,9) have higher alcohol% than quality levels 5,6,7, however, even quality 3,4 have higher alcohol% than 5,6,7. This shows that in some features(alcohol, pH, chlorides and fixed acidity) quality 3 and 4 contradict the 
trend and have characteristics similar to the higher qualities.

------

# Reflection

In  this exploration of white wine dataset, my main aim was to determine which 
features actually increase the quality of white wine. The alcohol%, density and 
chlorides turned out to be significant variables on simple boxplot analysis. Apart
from these, density variable turned out to be highly dependent on residual sugar 
and alcohol %. 
Most of the multivariate analysis plots were not very conclusive as quality level
3 and 4 displayed properties similar to higher quality levels in some features. 
Also, the dataset had very few samples of quality 3,4, 8 and 9. A huge number of 
samples belonged to level 6 and 7. 
This dataset is quite old(2009) and it should be updated with new samples of low 
and high qualities. Also we know that the quality of white wine can range from 0-10,
however our dataset only contains samples of quality 3-9.




